---
title: "Myleoid Erthrocyte Analysis"
author: "Kieran Campbell"
date: "30 January 2017"
output:
  html_document:
    highlight: tango
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE, cache = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, fig.width = 7, fig.height = 4,
                      warning = FALSE, message = FALSE)

library(scater)
library(tidyverse)

library(monocle)
library(dpt)
library(GGally)
library(viridis)
library(devtools)
library(cowplot)


devtools::load_all("~/oxford/mfa/mfa")

```

# Load data

```{r load-data}
raw <- read_csv("../data/wishbone_myleoid_erythrocyte_monocyte.txt.zip")
```

```{r convert-to-sceset}
gex <- select(raw, -tSNE1, -tSNE2, -Trajectory, -Branch)
pdata <- select(raw, tSNE1, tSNE2, Trajectory, Branch)

sample_names <- paste0("cell_", seq_len(nrow(gex)))
gex <- t(as.matrix(gex))
colnames(gex) <- sample_names

pdata <- as.data.frame(pdata)
rownames(pdata) <- sample_names

sce <- newSCESet(exprsData = gex, phenoData = AnnotatedDataFrame(pdata))
```


# Plots

```{r pca-plots}
plotPCA(sce, colour_by = "Trajectory", ncomponents = 3)
plotPCA(sce, colour_by = "Branch") + scale_fill_brewer(palette = "Set1")
```


# Analysis on a subset

```{r subset}
set.seed(123L)
sce_subset <- sce[, sample(ncol(sce), 2000)]

y <- t(exprs(sce_subset))

sce_subset <- plotPCA(sce_subset, colour_by = "Branch", return_SCESet = TRUE)
plotPCA(sce_subset, colour_by = "Trajectory")
```



```{r mfa}
set.seed(123L)
y_scaled <- scale(y)
# m <- mfa(y_scaled, iter = 1e5, thin = 50, pc_initialise = 2, 
#          b = 2, tau_c = 2, tau_eta = 1,
#         alpha = 1e-2, beta = 1,
#         alpha_chi = 5e3, beta_chi = 1)
# 
# print( m )

m <- mfa(y_scaled, iter = 1e5, thin = 50, pc_initialise = 2, 
         b = 2, tau_c = 1, tau_eta = 1,
        alpha = 2, beta = 1,
        alpha_chi = 5e3, beta_chi = 1)

print( m )
```

MCMC diagnostics:

```{r mcmc-diagnostics}
plot_mfa_trace(m)
plot_mfa_autocorr(m)
```

```{r mfa-summary}
mfa_summary <- summary(m)
print(mfa_summary)
```

```{r delete-me}
save(m, mfa_summary, file = "~/Desktop/deleteme.Rdata")
```

add this to SCESet:

```{r add-to-sceset}
for(col in names(mfa_summary)) pData(sce_subset)[[col]] <- NULL
pData(sce_subset) <- cbind(pData(sce_subset), mfa_summary)
```

and plots:

```{r plots, fig.width = 5, fig.height = 5}
# cowplot::plot_grid(plotPCA(sce_subset, colour_by = "branch"),
#           plotPCA(sce_subset, colour_by = "pseudotime"), nrow = 2)

plt1 <- plotPhenoData(sce_subset, aes_string(x = 'tSNE1', y = "tSNE2", color = "pseudotime"))
plt2 <- plotPhenoData(sce_subset, aes_string(x = 'tSNE1', y = "tSNE2", 
                            color = "branch"))


print(plt1)
print(plt2)
```

Proper expression plotting:

```{r proper-plot}
gex_tidy <- scale(y) %>% as_data_frame() %>% 
  dplyr::mutate(pseudotime = mfa_summary$pseudotime, branch = factor(mfa_summary$branch)) %>% 
  gather(gene, expression, -pseudotime, -branch)

cmap <- apply(m$traces$c_trace, 3, colMeans)
kmap <- apply(m$traces$k_trace, 3, colMeans)

pbranch1 <- colMeans(m$traces$gamma_trace - 1)

pex <- lapply(1:2, function(b) {
  b_expression <- sapply(1:m$G, function(g) {
    cmap[g,b] + kmap[g,b] * mfa_summary$pseudotime
  }) 
  colnames(b_expression) <- m$feature_names
  
  b_expression <- as_data_frame(b_expression) %>% 
    mutate(pseudotime = mfa_summary$pseudotime,
           branch = factor(b),
           pbranch1)
  gather(b_expression, gene, predicted_expression, -pseudotime, -branch, -pbranch1)
})

pex <- bind_rows(pex)

df <- inner_join(gex_tidy, pex, by = c("pseudotime", "branch", "gene"))
df <- arrange(df, pseudotime)

ggplot(df, aes(x = pseudotime, y = expression, color = branch)) +
  geom_point(alpha = 0.3) + facet_wrap(~ gene, scales = "free_y") + scale_color_brewer(palette = "Set1") +
  geom_line(aes(y = predicted_expression)) +
  scale_x_reverse()

```


Pi:

```{r calculate-pi}
wmap <- tabulate(round(colMeans(m$traces$gamma_trace)))
taumap <- colMeans(m$traces$tau_trace)
pi <- calculate_pi(y, cmap, kmap, mfa_summary$pseudotime, 
                   taumap, mean(m$traces$eta_trace),
                   1e-1, F, log(wmap / sum(wmap)), F)

sce_subset$pi <- pi[,1]
plotPCA(sce_subset, colour_by = "pi")

```



Figure for supplementary:

```{r fig-for-supp}
filter(df, gene %in% c("CD34", "CD11b")) %>% 
  ggplot(aes(x = pseudotime, y = expression)) +
    geom_point(aes(fill = 1 - pbranch1), alpha = 0.2, shape = 21) + 
    facet_wrap(~ gene, scales = "free_y") + 
    scale_fill_viridis(name = "p(branch 1)") +
  # scale_color_brewer(palette = "Set1", name = "Branch") +
    geom_line(aes(linetype = branch, y = predicted_expression), size = 1.5, alpha = 0.8) +
    scale_x_reverse() +
  xlab("Pseudotime") + ylab("Expression")

```


And save:

```{r save-fig}
ggsave("../figs/limits_of_chi.png", width=7, height=3)

```

```{r plot-gex, fig.width = 12, fig.height = 10}
gex_tidy <- y %>% as_data_frame() %>% 
  dplyr::mutate(pseudotime = mfa_summary$pseudotime, branch = factor(mfa_summary$branch)) %>% 
  gather(gene, expression, -pseudotime, -branch)

ggplot(gex_tidy, aes(x = pseudotime, y = expression, color = branch)) +
  geom_point(alpha = 0.3) + stat_smooth(se = FALSE, method = "lm") + facet_wrap(~ gene, scales = "free_y") + scale_color_brewer(palette = "Set1")

```

```{r plot-gex-errorbars}
ggplot(mfa_summary, aes(x = pseudotime, y = pseudotime, color = factor(branch), 
                        ymin = pseudotime_lower, ymax = pseudotime_upper)) + 
  geom_point() + geom_errorbar()

```



```{r chris-style-plot, fig.width = 12, fig.height = 10}
ggplot(gex_tidy, aes(x = pseudotime, y = gene, 
                    fill = expression)) + 
  geom_jitter(shape = 21, color = 'black') +
  facet_wrap(~ branch) + theme_classic() + viridis::scale_fill_viridis() 
```



## Plots for figure:

```{r plots-for-figure}
library(cowplot)
pd_df <- select(pData(sce_subset), tSNE1, tSNE2, pseudotime, branch) 

pst_plot <- ggplot(pd_df, aes(x = tSNE1, y = tSNE2, color = pseudotime)) +
  geom_point(alpha = 0.7) + scale_color_viridis(name = expression(paste("MAP ", t)))

branch_plot <- ggplot(pd_df, aes(x = tSNE1, y = tSNE2, color = branch)) +
  geom_point(alpha = 0.7) + 
  scale_color_brewer(palette = "Set1", name = expression(paste("MAP ", gamma)))

```

```{r plots-for-figure2}
chimap <- matrixStats::colMedians(m$traces$chi_trace)
chi_df <- data_frame(gene = featureNames(sce_subset), chi_inv = 1 / chimap)
# top <- dplyr::arrange(chi_df, desc(chi_inv)) %>% extract2("gene") %>% head(n = 20)
# bottom <- dplyr::arrange(chi_df, chi_inv) %>% extract2("gene") %>% head(n = 20)

chi_df_reduced <- dplyr::arrange(chi_df, desc(chi_inv))
chi_df_reduced$gene <- factor(chi_df_reduced$gene, levels = chi_df_reduced$gene)

breaks <- quantile(chi_df_reduced$chi_inv, probs = c(0.2, 0.8))
names(breaks) <- NULL
breaks <- round(breaks, 4)

chi_plot <- ggplot(chi_df_reduced, aes(x = gene, y = chi_inv)) +
  geom_bar(stat = 'identity') + 
  coord_flip(ylim = c(min(chi_df_reduced$chi_inv), max(chi_df_reduced$chi_inv))) +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(size = 9)) +
  ylab(expression(paste("[MAP ", chi[g] ,"]" ^ "-1")))+
  scale_y_continuous(breaks = breaks)

genes <- c("CD34", "CD235ab", "CD64")
gex <- t(exprs(sce_subset)[genes, ])

pd_tidy <- as_data_frame(cbind(pd_df, gex)) %>% 
  gather(gene, Expression, -(tSNE1:branch))

plot_gene <- function(g) {
  plt <- ggplot(dplyr::filter(pd_tidy, gene == g), aes(x = tSNE1, y = tSNE2, color = log(Expression + 1))) +
  geom_point() + scale_color_viridis(name = g) 
  if(g == "CD235ab") {
    plt <- plt + theme(
      legend.title = element_text(size = 9)
    )
  }
  return(plt)
}

expression_plots <- lapply(genes, plot_gene)

ex_plot <- plot_grid(plotlist = expression_plots, ncol = 1)


```

Combine into upper grid:

```{r upper-grid}
upper_grid <- plot_grid(
  pst_plot, branch_plot, nrow = 1, labels = c("A", "B")
)

```

# Comparison to alternative pseudotime algorithms


## plot PCA and find root cell

```{r plot-pca}
sce_subset <- plotPCA(sce_subset, return_SCESet = TRUE, colour_by = "Trajectory")
root_cell <- which.min(redDim(sce_subset)[,2])
pc2 <- redDim(sce_subset)[,2]

```


```{r fit-monocle}
to_sample_for_monocle <- sample(1:ncol(sce_subset), 1000, replace = FALSE)
monocle_pseudotime <- monocle_branch <- rep(NA, ncol(sce_subset))
monocle_root_cell <- which.min(sce_subset$Trajectory[to_sample_for_monocle])

cds <- toCellDataSet(sce_subset)# [, to_sample_for_monocle])
cds <- reduceDimension(cds, norm_method = "none")
cds <- orderCells(cds, num_paths = 2)
cds <- orderCells(cds, num_paths = 2, root_state = cds$State[root_cell])

monocle_pseudotime <- cds$Pseudotime # [to_sample_for_monocle] <- cds$Pseudotime
monocle_branch <- cds$State # [to_sample_for_monocle] <- cds$State

plot_cell_trajectory(cds, color_by = "Pseudotime")
```

## DPT

```{r dpt}
# gene_sds <- matrixStats::rowSds(exprs(s))
# gene_collapse_sd <- 1e-6 * gene_sds
# new_exprs <- apply(exprs(s), 2, function(x) rnorm(length(x), x, gene_collapse_sd))

to_sample_for_dpt <- sample(1:ncol(sce_subset), 2000, replace = FALSE)
dpt_pseudotime <- dpt_branch <- rep(NA, ncol(sce_subset))
dpt_root_cell <- which.min(sce_subset$Trajectory[to_sample_for_dpt])

ts <- Transitions(t(exprs(sce_subset))) #[, to_sample_for_dpt])))
pt <- dpt(ts, branching = TRUE, root = root_cell) 

dpt_pseudotime <- pt$DPT # [to_sample_for_dpt] <- pt$DPT
dpt_branch <- pt$Branch # [to_sample_for_dpt] <- pt$Branch
```



## Compare all

```{r compare-all}
df_compare <- data_frame(Wishbone = sce_subset$Trajectory,
                         mfa = sce_subset$pseudotime,
                         Monocle = monocle_pseudotime,
                         DPT = dpt_pseudotime,
                         PC2 = pc2)

df_tidy <- gather(df_compare, algorithm, value, -mfa) 
df_tidy$algorithm <- factor(df_tidy$algorithm, 
                            levels = c("PC2", "Wishbone", "Monocle", "DPT"))

pst_compare_plot <- ggplot(df_tidy, aes(x = mfa, y = value)) + 
  facet_wrap(~ algorithm, scales = "free_y", nrow = 1) +
  geom_point(alpha = 0.6) +
  xlab("MFA pseudotime") + ylab("Algorithm pseudotime")
```

Print correlations:

```{r print-cor}
cor(df_compare, use = 'complete')
```



Comparison of branch allocations:

```{r compare_branch_allocations}

## We're going to need to reduce the frequency of Monocle branches
bt <- table(cds$State)

infrequent_states <- bt < quantile(bt, probs = 0.7)
freq_stat <- plyr::mapvalues(cds$State, 
                             from = which(infrequent_states),
                             to = rep("Other", length(which(infrequent_states))))

branch_df <- data_frame(Wishbone = factor(sce_subset$Branch),
                        Monocle = factor(freq_stat), #cds$State,
                        DPT = factor(dpt_branch),
                        tSNE1 = sce_subset$tSNE1,
                        tSNE2 = sce_subset$tSNE2)

algorithms <- c("Wishbone", "Monocle", "DPT")

plot_bif <- function(alg) {
  plt <- ggplot(branch_df, aes_string(x = "tSNE1", y = "tSNE2", color = alg)) + 
    geom_point() + 
    cowplot::theme_cowplot() +
    theme(legend.title = element_text(size = 10, face = "bold"))
  if(alg == "Monocle") {
    plt <- plt + scale_color_discrete(name = paste(alg,"\nbranch"))
  } else {
    plt <- plt + scale_color_brewer(palette = "Set1", name = paste(alg,"\nbranch"))
  }
  return(plt)
}

bif_plots <- lapply(algorithms, plot_bif)

bottom_grid <- cowplot::plot_grid(plotlist = bif_plots, nrow = 1,
                   rel_widths = c(3,3,3.7))

```

Put it together for overall plot:

```{r overall-plot}

full_plot <- plot_grid(
  upper_grid,
  pst_compare_plot,
  bottom_grid,
  rel_heights = c(1.5,1,0.8), labels = c("", "C", "D"),
  ncol = 1
)

ggsave("../figs/masscyto_fig.png", full_plot, width = 10, height = 11)

```

Some diagnostics:

```{r diagnostics}
k0 <- colMeans(m$traces$k_trace[,,1])
k1 <- colMeans(m$traces$k_trace[,,2])
theta <- colMeans(m$traces$theta_trace)
chi <- colMeans(m$traces$chi_trace)

d <- data_frame(k0, k1, theta, chi, gene = m$feature_names) %>% 
  gather(variable, value, -gene) 

cowplot::plot_grid(
  filter(d, variable != "chi") %>% 
    ggplot(aes(x = gene, y = value, color = variable)) + geom_point() +
    theme(legend.position = 'bottom'),
  filter(d, variable == "chi") %>% 
    ggplot(aes(x = gene, y = value)) + geom_point(),
  ncol = 1)

```