---
title: "HSPC scRNA-seq analysis"
author: "Kieran Campbell"
date: "January 2017"
output:
  html_document:
    highlight: tango
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, fig.width = 7, fig.height = 4,
                      warning = FALSE, message = FALSE)


library(tidyverse)
library(scater)
library(viridis)
library(magrittr)
library(cowplot)

library(monocle)
library(dpt)
library(GGally)

devtools::load_all("~/oxford/mfa/mfa")
```

# Load data

Create an `SCESet`:

```{r make-sceset}
raw <- read_csv("../data/wishbone_mouse_marrow_scrnaseq.csv.gz")
tpm <- as.matrix(raw[,-1])

nc <- ncol(tpm)
wishbone_data <- data.frame(tpm[, (nc - 3):nc])
tpm <- tpm[, 1:(nc - 4)]

cellnames <- raw[[1]]
pd <- dplyr::mutate(wishbone_data, sample = cellnames)

rownames(pd) <- cellnames
tpm <- t(tpm)
colnames(tpm) <- cellnames

sce <- newSCESet(exprsData = tpm, phenoData = new("AnnotatedDataFrame", pd))
#is_exprs(sce) <- exprs(sce) > 0
#sce <- calculateQCMetrics(sce)
```

Let's subset to something reasonable:

```{r subset}
sc <- sce[rowMeans(exprs(sce) > 0) > 0.2, ] # expressed in at least 20% of cells
rm(sce)
```

# Plots

For paper?

```{r plot-pca}
plotPCA(sc, colour_by = "Branch") + 
  scale_fill_brewer(palette = "Set1", name = "Wishbone\nBranch")
```


# Inference

```{r gene-selection}
means <- rowMeans(exprs(sc))
vars <- matrixStats::rowVars(exprs(sc))
to_use <- vars > 5
```

```{r create-gex-matrix}
set.seed(123L)
# to_sample <- seq_len(ncol(sce)) 
to_sample <- sample(seq_len(ncol(sc)), 500)
s <- sc[to_use, to_sample]
X <- t(exprs(s))
rm(sc)
```

```{r inference}
set.seed(123L)
mrna <- mfa(scale(X), iter = 40000, thin = 20, pc_initialise = 2, 
         b = 2, tau_eta = 1e-2, tau_theta = 1e-2,
        tau_c = 1)
```

MCMC diagnostics:

```{r mcmc-diagnostics}
plot_mfa_trace(mrna)
plot_mfa_autocorr(mrna)
```

```{r mfa-summary}
mfa_summary <- summary(mrna)
print(mfa_summary)
```

add this to SCESet:

```{r add-to-sceset}
for(col in names(mfa_summary)) pData(s)[[col]] <- NULL
pData(s) <- cbind(pData(s), mfa_summary)
```

and plots:

```{r plots}
# ggplot(pData(s), aes(x = tSNE1, y = tSNE2, color = pseudotime)) + geom_point() +
#   scale_color_viridis()
# 
# ggplot(pData(s), aes(x = tSNE1, y = tSNE2, color = factor(branch), 
#                      size = -branch_certainty)) + 
#   geom_point() +
#   scale_color_brewer(palette = "Set1")
# 


plt1 <- plotPhenoData(s, aes_string(x = 'tSNE1', y = "tSNE2", color = "pseudotime"))
plt2 <- plotPhenoData(s, aes_string(x = 'tSNE1', y = "tSNE2", color = "branch"))
plt3 <- plotPhenoData(s, aes_string(x = 'tSNE1', y = "tSNE2", color = "branch_certainty"))

cowplot::plot_grid(plt1, plt2, plt3)
```

Plot bifurcating genes:

```{r bif-genes}
chi <- colMeans(mrna$traces$chi_trace)
chi_df <- data_frame(gene = featureNames(s), value = chi)

plotExpression(s, x = "pseudotime", colour_by = "branch",
               features = dplyr::arrange(chi_df, value) %>% magrittr::extract2("gene") %>% head(n=10),
               show_violin = FALSE, scales = 'free_y') + stat_smooth()
```


## Plots for figure:

```{r plots-for-figure}
library(cowplot)
pd_df <- select(pData(s), tSNE1, tSNE2, pseudotime, branch) 

pst_plot <- ggplot(pd_df, aes(x = tSNE1, y = tSNE2, color = pseudotime)) +
  geom_point(alpha = 0.7) + scale_color_viridis(name = expression(t))

branch_plot <- ggplot(pd_df, aes(x = tSNE1, y = tSNE2, color = branch)) +
  geom_point(alpha = 0.7) + 
  scale_color_brewer(palette = "Set1", name = expression(paste("MAP ", gamma)))

```

```{r plots-for-figure2}
chimap <- colMeans(mrna$traces$chi_trace)
chi_df <- data_frame(gene = featureNames(s), chi_inv = 1 / chimap)
top <- dplyr::arrange(chi_df, desc(chi_inv)) %>% extract2("gene") %>% head(n = 20)
bottom <- dplyr::arrange(chi_df, chi_inv) %>% extract2("gene") %>% head(n = 20)

chi_df_reduced <- dplyr::filter(chi_df, gene %in% c(top, bottom)) %>% 
  dplyr::arrange(desc(chi_inv))
chi_df_reduced$gene <- factor(chi_df_reduced$gene, levels = chi_df_reduced$gene)

chi_plot <- ggplot(chi_df_reduced, aes(x = gene, y = chi_inv)) +
  geom_bar(stat = 'identity') + coord_flip() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(size = 9)) +
  ylab(expression(paste("[MAP ", chi[g] ,"]" ^ "-1"))) 

genes <- c("ELANE", "CAR2", "RPL26")
gex <- t(exprs(s)[genes, ])

pd_tidy <- as_data_frame(cbind(pd_df, gex)) %>% 
  gather(gene, Expression, -(tSNE1:branch))

plot_gene <- function(g) {
  ggplot(dplyr::filter(pd_tidy, gene == g), aes(x = tSNE1, y = tSNE2, color = log(Expression + 1))) +
  geom_point() + scale_color_viridis(name = g) 
}

expression_plots <- lapply(genes, plot_gene)

ex_plot <- plot_grid(plotlist = expression_plots, ncol = 1)
```

Combine into upper grid:

```{r upper-grid}
upper_grid <- plot_grid(
  plot_grid(pst_plot, branch_plot, ncol = 1, labels = c("A", "B")),
  chi_plot,
  ex_plot,
  ncol = 3, labels = c("", "C", "D"), rel_widths = c(3,2.4,2.4)
)

```


# Comparison to other pseudotime algorithms

## plot PCA and find root cell

```{r plot-pca}
s <- plotPCA(s, return_SCESet = TRUE, colour_by = "Trajectory")
root_cell <- which.max(redDim(s)[,2])
pc2 <- redDim(s)[,2]
```


```{r fit-monocle}
cds <- toCellDataSet(s)
cds <- reduceDimension(cds, norm_method = "none")
cds <- orderCells(cds, num_paths = 2)
cds <- orderCells(cds, num_paths = 2, root_state = cds$State[root_cell])

plot_cell_trajectory(cds, color_by = "Pseudotime")
```

## DPT

```{r dpt}
# gene_sds <- matrixStats::rowSds(exprs(s))
# gene_collapse_sd <- 1e-6 * gene_sds
# new_exprs <- apply(exprs(s), 2, function(x) rnorm(length(x), x, gene_collapse_sd))

ts <- Transitions(t(exprs(s)))
pt <- dpt(ts, branching = TRUE, root = root_cell) 
```



## Compare all

```{r compare-all}
df_compare <- data_frame(Wishbone = s$Trajectory,
                         mfa = s$pseudotime,
                         Monocle = cds$Pseudotime,
                         DPT = pt$DPT,
                         PC2 = pc2)

df_tidy <- gather(df_compare, algorithm, value, -mfa) 
df_tidy$algorithm <- factor(df_tidy$algorithm, 
                            levels = c("PC2", "Wishbone", "Monocle", "DPT"))

pst_compare_plot <- ggplot(df_tidy, aes(x = mfa, y = value)) + 
  facet_wrap(~ algorithm, scales = "free_y", nrow = 1) +
  geom_point(alpha = 0.6) +
  xlab("MFA pseudotime") + ylab("Algorithm pseudotime")
```


Comparison of branch allocations:

```{r compare_branch_allocations}
branch_df <- data_frame(Wishbone = factor(s$Branch),
                        Monocle = cds$State,
                        DPT = pt$Branch,
                        tSNE1 = s$tSNE1,
                        tSNE2 = s$tSNE2)

algorithms <- c("Wishbone", "Monocle", "DPT")

plot_bif <- function(alg) {
  ggplot(branch_df, aes_string(x = "tSNE1", y = "tSNE2", color = alg)) + 
    geom_point() + 
    scale_color_brewer(palette = "Set1", name = paste(alg,"\nbranch")) +
    cowplot::theme_cowplot() +
    theme(legend.title = element_text(size = 10, face = "bold"))
}

bif_plots <- lapply(algorithms, plot_bif)

bottom_grid <- cowplot::plot_grid(plotlist = bif_plots, nrow = 1,
                   rel_widths = c(3,3,3.7))

```

Put it together for overall plot:

```{r overall-plot}

plot_grid(
  upper_grid,
  pst_compare_plot,
  bottom_grid,
  rel_heights = c(2.7,1,1), labels = c("", "E", "F"),
  ncol = 1
)

ggsave("../figs/scrnaseq_fig.png", width = 10, height = 12)

```


## Testing with larger number of components

```{r larger-components}
set.seed(123L)

mrnashrink <- mfa(scale(X), iter = 40000, thin = 20, pc_initialise = 2, 
         b = 5, tau_eta = 1, tau_theta = 1, prop_collapse = 0,
         alpha = 1, beta = 1, tau_c = 0.1, alpha_chi = 1000, beta_chi = 1, 
         w_alpha = 100)

mfa_summary <- summary(mrnashrink)
print(mfa_summary)

s$shrink_branch <- mfa_summary$branch
s$shrink_pst <- mfa_summary$pseudotime

plotPhenoData(s, aes(x = tSNE1, y = tSNE2, color = shrink_branch))
plotPhenoData(s, aes(x = tSNE1, y = tSNE2, color = shrink_pst))

```


## Shuffling of trunk

```{r shuffle}

sce_shuffle <- sc

exprs_branch_1 <- exprs(sce_shuffle[, sce_shuffle$Branch == 1])

exprs_branch_1_shuffled <- apply(exprs_branch_1, 1, 
                                 function(x) x[sample(seq_along(x))])

exprs(sce_shuffle)[, sce_shuffle$Branch == 1] <- 
  t(exprs_branch_1_shuffled)

message(varLabels(sce_shuffle))

cowplot::plot_grid(
  plotPCA(sce, colour_by = "Trajectory"),
  plotPCA(sce_shuffle, colour_by = "Trajectory"),
  labels = c("Original", "Branch 1 shuffled")
  )

# pct_dropout <- colMeans(exprs(sce) == 0) * 100
# qplot(pct_dropout) + xlab("% dropout")

plotTSNE(sce_shuffle[,sample(seq_len(ncol(sce_shuffle)), 800)], colour_by = "Trajectory", perplexity = 10)

plotTSNE(sc[,sample(seq_len(ncol(sc)), 800)], colour_by = "Trajectory", perplexity = 10)
```

